Understanding Implications of Dataset Choice for Feature Effect Estimation: A Simulation-Based Investigation through Error Decomposition

### 2025-02-14 paper/paper.tex
paper: add chapter on variance decomposition ablation study

\paragraph{Variance decomposition study.} The variance of the feature effects estimated with the procedure described in
the main experiment \textsc{Section~\ref{sec:experiment-main}} consists of two
components: the model variance and the variance due to the Monte Carlo
integration. To disentangle these two sources of variance, we aim to
additionally estimate the Monte Carlo variance using the estimator in
equation~(\ref{eq:mc-variance-estimator-model}). This allows us to substract it
from the total variance estimated in the main experiment and obtain the model
variance, giving more nuanced insights into the implications of the data choice
on feature effect estimation. To obtain this estimate, we maintain the set-up
of the main experiment but for each trained model $\hat f^{(m)}$ (plus five
models from CV $\hat f^{(m,1)} - \hat f^{(m,5)}$), we draw $K=30$ new training
and validation sets from the DGP and estimate the feature effects on these sets
plus in a cross-validated manner with the already fitted models. This allows us
to estimate the Monte Carlo variance as in
equation~(\ref{eq:mc-variance-estimator-model}). Since this study is
computationally more expensive, we do this only for XGBoost (OT and OF) with
the same datasets and sizes as in the main experiment.

@@

Generally, we observe that the feature effect error is in the most cases highest when
the effect is estimated on validation data. For ALE, this observation is more pronounced
than for PDP. This increase in error is mainly due to an increase in variance (see e.g.,  % chktex 13
\textbf{Fig.\@~\ref{fig:pdp-results-snc-gam-ot}} or \textbf{Fig.\@~\ref{fig:ale-results-snc-xgboost-of}}).
For many models, the feature effect error is lowest when estimated in a cross-validated manner.
This is slightly more pronounced for overfitting models (OF) than for optimally tuned models (OT),
and is mainly due to a decrease in variance (see e.g., \textbf{Fig.\@~\ref{fig:pdp-results-snc-gam-of}}
and \textbf{Fig.\@~\ref{fig:pdp-results-snc-gam-ot}}). Since the variance composes of the model variance
and the Monte Carlo variance, both could be reduced by cross-validation. Monte Carlo variance
may decrease as we have slightly more samples overall than in the training set. Conversely,
the model variance may decrease as it is averaged out through the cross-validation procedure,
which may especially improve overfitting models that have a higher model variance.
We investigate these hypotheses in an ablation study in \textsc{Section~\ref{sec:results-variance-decomposition}}.

@@

Further results on Friedman1 and Feynman I.29.16, as well as for the dummy features, largely confirm
the described findings. For details, we refer to \textsc{Appendix~\ref{app:additional-results}}. We only
observe one case where the error of ALE is substantially larger when estimated on training data
due to a high bias. This occurs for the XGBoost\_OF model on the Friedman1 dataset, which we further
look into in \textsc{Appendix~\ref{app:main-study-further-investigation}}.

@@

\dots

Finally, we also observe that the Monte Carlo variance --- as expected --- reduced with larger dataset size.
We investigate this effect further in the next ablation study.

@@

\begin{figure*}[h!]
    \centering
    \begin{subfigure}[b]{\textwidth}
        \centering
        \includegraphics[width=0.9\textwidth]{img/model_performance_SimpleNormalCorrelated.png}
        \caption{Model performances on SimpleNormalCorrelated}
    \end{subfigure}
    \\[10pt]
    \vfill
    \begin{subfigure}[b]{\textwidth}
        \centering
        \includegraphics[width=0.9\textwidth]{img/model_performance_Friedman1.png}
        \caption{Model performances on Friedman1}
    \end{subfigure}
    \\[10pt]
    \vfill
    \begin{subfigure}[b]{\textwidth}
        \centering
        \includegraphics[width=0.9\textwidth]{img/model_performance_FeynmanI.29.16.png}
        \caption{Model performances on Feynman I.29.16}
    \end{subfigure}
    \caption{R2-scores of the models evaluated on training samples (left) and 10000 holdout test samples (right). Each boxplot aggregates 30 repetitions with differing seeds for the data generation (cf. \textsc{Section~\ref{sec:experiment-main}}).}
    \label{fig:model-performance}  % chktex 24
\end{figure*}

@@

@@

@@

>>>>

\paragraph{Variance decomposition study.} The variance of the feature effects estimated with the procedure described in
the main experiment \textsc{Section~\ref{sec:experiment-main}} consists of two
components: the model variance and the variance due to the Monte Carlo
integration. To disentangle these two sources of variance, we aim to
additionally estimate the Monte Carlo variance using the estimator in
equation~(\ref{eq:mc-variance-estimator-model}). This allows us to substract it
from the total variance estimated in the main experiment and obtain the model
variance, giving more nuanced insights into the implications of the data choice
on feature effect estimation. To obtain this estimate, we maintain the set-up
of the main experiment but for each trained model $\hat f^{(m)}$ (plus five
models from CV $\hat f^{(m,1)} - \hat f^{(m,5)}$), we draw $K=30$ new training
and validation sets from the DGP and estimate the feature effects on these sets
plus in a cross-validated manner with the already fitted models. This allows us
to estimate the Monte Carlo variance as in
equation~(\ref{eq:mc-variance-estimator-model}). Since this study is
computationally more expensive, we do this only for XGBoost (OT and OF), which
is arguably the most practically relevant model in our set-up. We use the same
datasets and sizes as in the main experiment.

@@

Generally, we observe that the feature effect error is in the most cases highest when
the effect is estimated on validation data. For ALE, this observation is more pronounced
than for PDP. This increase in error is mainly due to an increase in variance (see e.g.,  % chktex 13
\textbf{Fig.\@~\ref{fig:pdp-results-snc-gam-ot}} or \textbf{Fig.\@~\ref{fig:ale-results-snc-xgboost-of}}).
For many models, the feature effect error is lowest when estimated in a cross-validated manner.
This is slightly more pronounced for overfitting models (OF) than for optimally tuned models (OT),
and is mainly due to a decrease in variance (see e.g., \textbf{Fig.\@~\ref{fig:pdp-results-snc-gam-of}}
and \textbf{Fig.\@~\ref{fig:pdp-results-snc-gam-ot}}). Since the variance composes of the model variance
and the Monte Carlo variance, both could be reduced by cross-validation. Monte Carlo variance
may decrease as we have slightly more samples overall than in the training set. Conversely,
the model variance may decrease as it is averaged out through the cross-validation procedure,
which may especially improve overfitting models that have a higher model variance.
We further investigate these hypotheses in \textsc{Section~\ref{sec:results-variance-decomposition}}.

@@

Further results on Friedman1 and Feynman I.29.16, as well as for the dummy features, largely confirm
the described findings. For details, we refer to \textsc{Appendix~\ref{app:additional-results}}. We only
observe one case where the error of ALE is substantially larger when estimated on training data
due to a high bias. This occurs for the XGBoost\_OF model on the Friedman1 dataset, which we further
look into in \textsc{Appendix~\ref{app:main-study-further-investigation}}. Also note that the we observe
no substantial differences between scenarios with correlated features (SimpleNormalCorrelated) and
independent features (Friedman1, Feynman I.29.16).

@@

We use the variance results from the main experiment and the Monte Carlo variance estimated in the ablation study
to show the decomposition of the feature effect variance into model variance and Monte Carlo variance. The results
for the Friedman1 dataset are given in \textbf{Fig.\@~\ref{fig:pdp-variance-decomp}} for PDP and
\textbf{Fig.\@~\ref{fig:ale-variance-decomp}} for ALE. The full results including the other datasets are  % chktex 13
provided in \textsc{Appendix~\ref{app:additional-variance-decomposition}}.

We observe that the Monte Carlo variance is generally highest when the feature effects are estimated on the smaller
validation set, which aligns with our expectations.
Computing the feature effects in a cross-validated manner leads to the lowest Monte Carlo variance with a substantial
reduction compared to estimation solely on training or validation data.

In addition, we observe that while the model variance is naturally nearly the same for training and validation data,
it is also reduced in many cases when estimating the feature effects in a
cross-validated manner. This is especially pronounced for the overfitting models (OF) and is best visible for PDP
(\textbf{Fig.\@~\ref{fig:pdp-variance-decomp}}). These observations are in line with our previous hypothesis that
reduces both model variance by averaging over different models as well as Monte Carlo variance by averaging over
different samples. This is especially beneficial for overfitting models, which have a higher model variance.

\begin{figure}[htbp]
    \centering
    \begin{subfigure}[b]{0.75\textwidth}
        \includegraphics[width=\textwidth]{img/variance_decomposition_pdp_XGBoost_OF.png}
        \caption{XGBoost\_OF PDP on Friedman1}
    \end{subfigure}
    \\[10pt]
    \vfill
    \begin{subfigure}[b]{0.75\textwidth}
        \includegraphics[width=\textwidth]{img/variance_decomposition_pdp_XGBoost_OT.png}
        \caption{XGBoost\_OT PDP on Friedman1}
    \end{subfigure}
    \caption{Variance decomposition results for PDP on Friedman1 dataset. Variance is taken from the
    main experiment, Monte Carlo variance is estimated in the ablation study. The model variance is
    obtained by subtracting the Monte Carlo variance from the total variance. The second linear feature
    ($x_5$) and dummy features ($x_6$, $x_7$) are omitted for better readability. Results for XGBoost\_OT
    on 8000 training samples were not obtained due to excessive computation time.}
    \label{fig:pdp-variance-decomp}  % chktex 24
\end{figure}

While the variance due to MC integration is only a minor component of the total variance for PDP, it is the
dominant component for ALE. Only for feature effect estimation via cross-validation, the Monte Carlo variance  % chktex 13
is again reduced to a negligible level. This suggests that ALE is more sensitive to the dataset size than PDP,
which we further investigate in \textsc{Section~\ref{sec:results-dataset-size}}. We also suspect a certain
amount of inaccuracy in the Monte Carlo variance estimation for ALE, which (a) leads to differences in the
model variance between training and validation set and (b) causes the model variance to be negative in some cases
(see \textbf{Fig.\@\ref{fig:ale-variance-decomp}}).
Naturally, this is not possible, and can most likely be attributed to numerical inaccuracies in the MC variance estimation.

\begin{figure}[htbp]
    \centering
    \begin{subfigure}[b]{0.75\textwidth}
        \includegraphics[width=\textwidth]{img/variance_decomposition_ale_XGBoost_OF.png}
        \caption{XGBoost\_OF ALE on Friedman1}
    \end{subfigure}
    \\[10pt]
    \vfill
    \begin{subfigure}[b]{0.75\textwidth}
        \includegraphics[width=\textwidth]{img/variance_decomposition_ale_XGBoost_OT.png}
        \caption{XGBoost\_OT ALE on Friedman1}
    \end{subfigure}
    \caption{Variance decomposition results for ALE on Friedman1 dataset. Variance is taken from the
    main experiment, Monte Carlo variance is estimated in the ablation study. The model variance is
    obtained by subtracting the Monte Carlo variance from the total variance. The second linear feature
    ($x_5$) and dummy features ($x_6$, $x_7$) are omitted for better readability. Results for XGBoost\_OT
    on 8000 training samples were not obtained due to excessive computation time. Negative variances are
    due to numerical inaccuracies, marked by the hatch pattern.}
    \label{fig:ale-variance-decomp}  % chktex 24
\end{figure}

For both PDP and ALE, we find that the Monte Carlo variance, as expected, reduces with larger dataset size.
This becomes clear both when comparing training and validation data and when comparing the two dataset sizes
(1000 and 8000 samples). To analyze this not only for these few dataset sizes, we dedicate a separate
ablation study to investigate this further in the next section.

@@

\begin{figure*}[h!]
    \centering
    \begin{subfigure}[b]{\textwidth}
        \centering
        \includegraphics[width=0.9\textwidth]{img/model_performance_SimpleNormalCorrelated.png}
        \caption{Model performances on SimpleNormalCorrelated}
    \end{subfigure}
    \\[10pt]
    \vfill
    \begin{subfigure}[b]{\textwidth}
        \centering
        \includegraphics[width=0.9\textwidth]{img/model_performance_Friedman1.png}
        \caption{Model performances on Friedman1}
    \end{subfigure}
    \\[10pt]
    \vfill
    \begin{subfigure}[b]{\textwidth}
        \centering
        \includegraphics[width=0.9\textwidth]{img/model_performance_FeynmanI.29.16.png}
        \caption{Model performances on Feynman I.29.16}
    \end{subfigure}
    \caption{R2-scores of the models evaluated on training samples (left) and 10000 holdout test samples (right). Each boxplot aggregates 30 repetitions with differing seeds for the data generation (cf.\ \textsc{Section~\ref{sec:experiment-main}}).}
    \label{fig:model-performance}  % chktex 24
\end{figure*}

@@

\clearpage
\section{Additional Results for Ablation Study: Variance Decomposition}\label{app:additional-variance-decomposition}

\subsection{SimpleNormalCorrelated Dataset}
\begin{figure}[htbp]
    \centering
    \begin{subfigure}[b]{0.75\textwidth}
        \includegraphics[width=\textwidth]{img/SNC-all/variance_decomposition_pdp_XGBoost_OF.png}
        \caption{XGBoost\_OF PDP on SimpleNormalCorrelated}
    \end{subfigure}
    \\[10pt]
    \vfill
    \begin{subfigure}[b]{0.75\textwidth}
        \includegraphics[width=\textwidth]{img/SNC-all/variance_decomposition_pdp_XGBoost_OT.png}
        \caption{XGBoost\_OT PDP on SimpleNormalCorrelated}
    \end{subfigure}
    \caption{Variance decomposition results for PDP on SimpleNormalCorrelated dataset. Variance is taken from the
    main experiment, Monte Carlo variance is estimated in the ablation study. The model variance is
    obtained by subtracting the Monte Carlo variance from the total variance. Results for XGBoost\_OT
    on 8000 training samples were not obtained due to excessive computation time.}
    \label{fig:pdp-variance-decomp-snc}  % chktex 24
\end{figure}

\begin{figure}[htbp]
    \centering
    \begin{subfigure}[b]{0.75\textwidth}
        \includegraphics[width=\textwidth]{img/SNC-all/variance_decomposition_ale_XGBoost_OF.png}
        \caption{XGBoost\_OF ALE on SimpleNormalCorrelated}
    \end{subfigure}
    \\[10pt]
    \vfill
    \begin{subfigure}[b]{0.75\textwidth}
        \includegraphics[width=\textwidth]{img/SNC-all/variance_decomposition_ale_XGBoost_OT.png}
        \caption{XGBoost\_OT ALE on SimpleNormalCorrelated}
    \end{subfigure}
    \caption{Variance decomposition results for PDP on SimpleNormalCorrelated dataset. Variance is taken from the
    main experiment, Monte Carlo variance is estimated in the ablation study. The model variance is
    obtained by subtracting the Monte Carlo variance from the total variance. Results for XGBoost\_OT
    on 8000 training samples were not obtained due to excessive computation time. Negative variances are
    due to numerical inaccuracies, marked by the hatch pattern.}
    \label{fig:ale-variance-decomp-snc}  % chktex 24
\end{figure}

\begin{figure}[htbp]
    \centering
    \begin{subfigure}[b]{\textwidth}
        \includegraphics[width=\textwidth]{img/Friedman1-all/variance_decomposition_pdp_XGBoost_OF.png}
        \caption{XGBoost\_OF PDP on Friedman1}
    \end{subfigure}
    \\[10pt]
    \vfill
    \begin{subfigure}[b]{\textwidth}
        \includegraphics[width=\textwidth]{img/Friedman1-all/variance_decomposition_pdp_XGBoost_OT.png}
        \caption{XGBoost\_OT PDP on Friedman1}
    \end{subfigure}
    \caption{Variance decomposition results for PDP on Friedman1 dataset. Variance is taken from the
    main experiment, Monte Carlo variance is estimated in the ablation study. The model variance is
    obtained by subtracting the Monte Carlo variance from the total variance. Results for XGBoost\_OT
    on 8000 training samples were not obtained due to excessive computation time.}
    \label{fig:pdp-variance-decomp-friedman1}  % chktex 24
\end{figure}

\begin{figure}[htbp]
    \centering
    \begin{subfigure}[b]{\textwidth}
        \includegraphics[width=\textwidth]{img/Friedman1-all/variance_decomposition_ale_XGBoost_OF.png}
        \caption{XGBoost\_OF ALE on Friedman1}
    \end{subfigure}
    \\[10pt]
    \vfill
    \begin{subfigure}[b]{\textwidth}
        \includegraphics[width=\textwidth]{img/Friedman1-all/variance_decomposition_ale_XGBoost_OT.png}
        \caption{XGBoost\_OT ALE on Friedman1}
    \end{subfigure}
    \caption{Variance decomposition results for PDP on Friedman1 dataset. Variance is taken from the
    main experiment, Monte Carlo variance is estimated in the ablation study. The model variance is
    obtained by subtracting the Monte Carlo variance from the total variance. Results for XGBoost\_OT
    on 8000 training samples were not obtained due to excessive computation time. Negative variances are
    due to numerical inaccuracies, marked by the hatch pattern.}
    \label{fig:ale-variance-decomp-friedman1}  % chktex 24
\end{figure}

\begin{figure}[htbp]
    \centering
    \begin{subfigure}[b]{0.9\textwidth}
        \includegraphics[width=\textwidth]{img/FeynmanI.29.16-all/variance_decomposition_pdp_XGBoost_OF.png}
        \caption{XGBoost\_OF PDP on FeynmanI.29.16}
    \end{subfigure}
    \\[10pt]
    \vfill
    \begin{subfigure}[b]{0.9\textwidth}
        \includegraphics[width=\textwidth]{img/FeynmanI.29.16-all/variance_decomposition_pdp_XGBoost_OT.png}
        \caption{XGBoost\_OT PDP on FeynmanI.29.16}
    \end{subfigure}
    \caption{Variance decomposition results for PDP on FeynmanI.29.16 dataset. Variance is taken from the
    main experiment, Monte Carlo variance is estimated in the ablation study. The model variance is
    obtained by subtracting the Monte Carlo variance from the total variance. Results for XGBoost\_OT
    on 8000 training samples were not obtained due to excessive computation time.}
    \label{fig:pdp-variance-decomp-feynman}  % chktex 24
\end{figure}

\begin{figure}[htbp]
    \centering
    \begin{subfigure}[b]{0.9\textwidth}
        \includegraphics[width=\textwidth]{img/FeynmanI.29.16-all/variance_decomposition_ale_XGBoost_OF.png}
        \caption{XGBoost\_OF ALE on FeynmanI.29.16}
    \end{subfigure}
    \\[10pt]
    \vfill
    \begin{subfigure}[b]{0.9\textwidth}
        \includegraphics[width=\textwidth]{img/FeynmanI.29.16-all/variance_decomposition_ale_XGBoost_OT.png}
        \caption{XGBoost\_OT ALE on FeynmanI.29.16}
    \end{subfigure}
    \caption{Variance decomposition results for PDP on FeynmanI.29.16 dataset. Variance is taken from the
    main experiment, Monte Carlo variance is estimated in the ablation study. The model variance is
    obtained by subtracting the Monte Carlo variance from the total variance. Results for XGBoost\_OT
    on 8000 training samples were not obtained due to excessive computation time. Negative variances are
    due to numerical inaccuracies, marked by the hatch pattern.}
    \label{fig:ale-variance-decomp-feynman}  % chktex 24
\end{figure}

@@

\subsection{Aggregated Results}

@@

\clearpage
\subsection{Results per Grid Point}
